---
title: "Exporting custom columns created in scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting custom columns created in scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What this is for

A scenario function can attach new columns to the synthetic population
(`synthpop$pop`). By default those columns are **dropped** before the
lifecourse is written to disk, so they never reach the summaries or tables.

This vignette explains the mechanism that lets a user-created column survive
to the lifecourse output **and** be aggregated into the summary files, with no
changes to the model's source code. The mechanism is purely **naming
convention** plus one entry in the design YAML.

## The two requirements

For a scenario-created column to appear in the summaries, it must satisfy
**both** of the following:

1. **Be listed in `cols_for_output`** in the design YAML (so it survives the
   lifecourse pruning step), *or* be named with a recognised suffix that is
   kept automatically (`_prvl`, `_contd`, `_costs`, `_dgns`, `_mrtl`, or a
   `cms_` prefix).
2. **Use a recognised suffix** so the export step knows how to aggregate it.

The recognised suffixes and their treatment are:

| Suffix    | Interpreted as        | Aggregation                              | Output folder(s)                                   |
|-----------|-----------------------|------------------------------------------|----------------------------------------------------|
| `_prvl`   | duration counter      | prevalence `SUM(wt)` where `col > 0` **and** incidence `SUM(wt)` where `col = 1` | `summaries/prvl_scaled_up/`, `summaries/incd_scaled_up/` (+ `_esp`) |
| `_contd`  | continuous quantity   | population-weighted mean                  | `summaries/contd_scaled_up/`, `summaries/contd_esp/` |
| `_costs`  | monetary cost         | `SUM(col * wt)`                           | `summaries/costs_scaled_up/`, `summaries/costs_esp/` |

`_prvl` and `_costs` ride on the existing exporters; `_contd` is exported by a
dedicated step that you opt into with `type = c(..., "contd")` in
`export_summaries()`.

## How `_prvl` yields both prevalence and incidence

The disease prevalence columns in the model are not 0/1 flags — they are
**duration counters**: the value is `1` in the year the condition first
appears (the incidence event), then `2`, `3`, … for each subsequent year the
person carries the condition. The prevalence exporter counts everyone with
`col > 0`; the incidence exporter counts only `col = 1`. A single column drives
both.

To build such a counter from a 0/1 event vector, use `carry_forward_incr()`,
the same helper the model uses internally:

```{r}
# inside a scenario function, `synthpop$pop` is keyed by (pid, year)
synthpop$pop[, my_event_prvl := carry_forward_incr(
  as.integer(some_condition),  # 1 in the event year, 0 otherwise
  pid_mrk,                     # per-simulant marker (already present)
  recur = FALSE,               # FALSE = once-only (e.g. CHD); TRUE = can recur
  y = 1L,
  byref = TRUE
)]
```

If you only care about prevalence and not incidence timing, a plain
non-decreasing indicator also works — anything `> 0` counts as prevalent and
anything `= 1` counts as incident in that year.

## A worked example

The example below defines a primary-prevention scenario that, from 2014
onwards, reduces systolic blood pressure (SBP) by 10% and records three custom
columns:

* `sbp_intervention_prvl` — a flag for "is currently exposed to the
  intervention", as a duration counter so it yields both a prevalence and an
  incidence summary;
* `sbp_excess_contd` — a continuous quantity (mmHg of SBP above 130), summarised
  as a population-weighted mean;
* `sbp_intervention_costs` — a per-person-year monetary cost of delivering the
  intervention.

### Step 1 — declare the columns in the design YAML

`_prvl`, `_contd` and `_costs` columns are kept automatically by the pruning
step, so strictly you do not need to list them. Listing them anyway is good
documentation and is **required** for any custom column that does *not* use a
recognised suffix:

```yaml
cols_for_output:
- pid
- pid_mrk
- year
- mc
- age
- agegrp
- sex
- wt
- wt_esp
- sbp_intervention_prvl    # optional for suffixed cols, but explicit is nice
- sbp_excess_contd
- sbp_intervention_costs
```

### Step 2 — create the columns in the scenario function

```{r}
source("./global.R")
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

IMPACTncd$
  del_logs()$
  del_outputs()$
  run(1:2, multicore = TRUE, "sc0")   # baseline

IMPACTncd$update_primary_prevention_scn(
  function(synthpop) {
    # the intervention itself: a 10% SBP reduction from 2014 onwards
    synthpop$pop[year > 2013, SBP_curr_xps := SBP_curr_xps * 0.9]

    # (1) _prvl: duration counter -> drives prvl AND incd summaries
    synthpop$pop[, sbp_intervention_prvl := carry_forward_incr(
      as.integer(year > 2013),
      pid_mrk,
      recur = FALSE,
      y = 1L,
      byref = TRUE
    )]

    # (2) _contd: continuous -> population-weighted mean
    synthpop$pop[, sbp_excess_contd := pmax(0, SBP_curr_xps - 130)]

    # (3) _costs: monetary -> SUM(col * wt); feeds net_costs downstream
    synthpop$pop[, sbp_intervention_costs := fifelse(year > 2013, 500, 0)]

    NULL
  }
)

IMPACTncd$
  run(1:2, multicore = TRUE, "sc1")   # intervention scenario
```

### Step 3 — export summaries, including the continuous one

`prvl`, `incd` and `costs` are standard types and pick up the custom columns
with no extra work. Add `"contd"` to also export the weighted-mean summary for
`_contd` columns.

```{r}
IMPACTncd$export_summaries(
  multicore = TRUE,
  type = c(
    "le", "hle", "dis_char",
    "prvl", "incd", "dis_mrtl", "mrtl",
    "all_cause_mrtl_by_dis", "cms",
    "qalys", "costs",
    "contd"            # <- exports summaries/contd_scaled_up/ etc.
  )
)
```

## What the outputs look like

Each summary file is one Parquet dataset per Monte Carlo iteration, stratified
by `strata_for_output` (by default `scenario, year, agegrp, sex`) plus `mc` and
a `popsize` column.

The idiomatic way to read a summary in this project is `arrow::open_dataset()`
on the summary *folder*, which combines all Monte Carlo files at once.

```{r}
library(arrow)
library(data.table)

smr <- file.path(IMPACTncd$design$sim_prm$output_dir, "summaries")

# (1) prevalence of the intervention flag
prvl <- as.data.table(open_dataset(file.path(smr, "prvl_scaled_up")))
prvl[, .(mc, scenario, year, agegrp, sex, popsize, sbp_intervention_prvl)]

# the SAME _prvl column also produced an incidence column (col == 1)
incd <- as.data.table(open_dataset(file.path(smr, "incd_scaled_up")))
incd[, .(mc, scenario, year, agegrp, sex, popsize, sbp_intervention_incd)]

# (2) weighted-mean of the continuous column
contd <- as.data.table(open_dataset(file.path(smr, "contd_scaled_up")))
contd[, .(mc, scenario, year, agegrp, sex, popsize, sbp_excess_contd)]

# (3) summed cost column, alongside the model's built-in cost columns
costs <- as.data.table(open_dataset(file.path(smr, "costs_scaled_up")))
costs[, .(mc, scenario, year, agegrp, sex, sbp_intervention_costs)]
```

Notes on the column names:

* The `_prvl` column keeps its name in the prevalence file
  (`sbp_intervention_prvl`) and is renamed `…_incd` in the incidence file
  (`sbp_intervention_incd`).
* The `_contd` column keeps its name and holds the weighted mean.
* The `_costs` column keeps its name and holds the weighted sum.

## Downstream tables (`auxil/process_out.R`)

The post-processing script `auxil/process_out.R` turns the per-Monte-Carlo
summaries into quantile tables. It selects columns by **suffix**, so most
custom columns are tabulated automatically:

* `_prvl` columns feed the `prvl`, `prvl_change` and `cypp` (case-years
  prevented/postponed) tables.
* the derived `_incd` columns feed `incd`, `incd_change` and `cpp`.
* `_costs` columns feed `costs` and `net_costs`.
* `_contd` columns feed the `contd` table (a population-weighted mean,
  re-weighted by `popsize` when collapsing strata).

You do not need to edit `process_out.R` to add a new custom column — naming it
with a recognised suffix is enough.

### Costs, net costs and economic evaluation

The `net_costs` output is the difference versus the comparator scenario (by
default `sc0`). Because it discovers all `_costs$` columns, your custom cost
appears in net-cost tables without further configuration.

Incremental cost-effectiveness ratios (ICER) and net monetary benefit (NMB) are
also produced directly by the `Simulation$export_tables()` method (the class
method that supersedes `process_out.R`). It writes
`cost-effectiveness by <strata> (<perspective>-<scale>) (not standardised).csv`
files containing the cumulative incremental QALYs (`dQALYs_cuml`) and costs
(`dCosts_cuml`), the `ICER`, and the `NMB` at each willingness-to-pay threshold,
at each discount level (see below), for both QALY scales (EQ5D5L and HUI3) and
from two perspectives:

* **societal** — uses `cvd_total_costs` (direct + productivity + informal care);
* **healthcare** — uses `cvd_direct_costs` (treatment costs only).

The willingness-to-pay thresholds are set with the `wtp` argument
(default `c(5e6, 7.5e6, 1e7)` ¥/QALY); each becomes its own `type` value
(`NMB_at_wtp_5000000`, …). See `?Simulation` / `export_tables()` for the full
argument list.

#### Discounting

`cost_discount_rate` and `qaly_discount_rate` take *vectors* of annual rates
(percent), defaulting to `c(0, 2)` for both. The two are paired element-wise
into discount **levels** (a length-1 rate is recycled against a longer one), so
the defaults give two levels: undiscounted, and 2% on costs and QALYs alike.
Differential rates are set by passing equal-length vectors, e.g.
`qaly_discount_rate = c(0, 1.5)` with `cost_discount_rate = c(0, 4)`.

Every level is reported in **all** the affected tables — `QALYs by …`,
`net QALYs by …`, `costs by …`, `net costs by …` *and*
`cost-effectiveness by …` — not just the cost-effectiveness ones. The file set
is unchanged: each of those tables gains a `discount` column and carries one
block of rows per level, labelled `"0%"`, `"2%"`, or
`"QALYs 1.5%/costs 4%"` when a level's two rates differ. Filter on `discount`
to pick a level; the `0%` rows are the undiscounted figures.

Present values are computed as
`PV = FV / (1 + rate/100)^max(0, year - discount_from_year)`, applied to each
year's flow *before* the cumulative columns are accumulated, so
`QALYs_cuml` / `costs_cuml` / `dQALYs_cuml` / `dCosts_cuml` are sums of present
values. `discount_from_year` defaults to `baseline_year_for_change_outputs`, and
years at or before it are undiscounted.

#### Discount levels and WTP thresholds

Two vectors meet in the cost-effectiveness tables, and they combine by
*different* rules — worth knowing before you read one:

| Pair | Rule | With `c(0, 2)` rates and 3 thresholds |
|---|---|---|
| `qaly_discount_rate` × `cost_discount_rate` | **paired** element-wise | 2 levels (not 4) |
| discount level × `wtp` | **crossed** | 2 × 3 = 6 NMB rows per stratum cell |

The rates are paired because a QALY rate and a cost rate together define *one*
analysis — that pairing is what lets `c(0, 1.5)` / `c(0, 4)` express "undiscounted"
and "1.5% QALYs, 4% costs" as two scenarios rather than a 2×2 grid you would have
to subset. `wtp` is an independent sensitivity axis, so every threshold is
evaluated at every discount level.

Both axes end up as *rows* (the tables are long), via different columns:
`discount` tags the level, and `type` carries the metric — so filter on **both**
to isolate a single series:

```r
dt <- fread("cost-effectiveness by year (societal-EQ5D5L) (not standardised).csv")
dt[discount == "2%" & type == "NMB_at_wtp_5000000"]
```

Row counts scale as
`scenarios × strata × levels × (3 + length(wtp))`, so adding a discount level
multiplies rows while leaving the *file* set untouched.

Two consequences of the crossing:

* **A threshold is never discounted.** It is a ratio of present values
  (¥ per QALY), so it multiplies the already-discounted `dQALYs_cuml`:
  `NMB = wtp × dQALYs_cuml − dCosts_cuml` is a present value at that level's
  rates. Under differential rates a single NMB figure therefore mixes both
  (QALYs at one rate, costs at the other) — which is exactly what the
  spelled-out `"QALYs 1.5%/costs 4%"` label is recording.
* **`ICER` does not vary with `wtp`.** It is `dCosts_cuml / dQALYs_cuml`, so it
  appears once per discount level, not once per threshold (it *does* vary with
  the discount level, since both parts are discounted). Its rows are absent
  wherever `dQALYs_cuml` is 0 and the ratio is undefined — typically the years
  before an intervention starts — so an ICER series is shorter than the NMB
  series beside it.

**Where do your custom `_costs` columns count?** Custom cost columns are
**always** added to the **societal** perspective. By default they are *not*
included in the healthcare perspective (which is meant to capture direct
treatment costs only). If some of your custom costs *are* healthcare costs,
pass their column names to `custom_costs_in_healthcare` in `export_tables()` and
they are added to the healthcare perspective as well, e.g.
`custom_costs_in_healthcare = c("screening_costs", "drug_costs")`. For example,
an intervention-delivery cost paid by the health system should be listed here,
whereas a productivity-loss or out-of-pocket cost should be left in the
societal perspective only. (Passing `TRUE` includes *all* custom cost columns
and `NULL`/`FALSE` includes none, for backward compatibility.)

If you need a bespoke calculation, you can still derive ICER and NMB yourself
from the `net_costs` and `net_qalys` tables in your own analysis scripts.

## Tips and gotchas

* **Keep the column present in every scenario.** If a column exists in `sc1`
  but not in `sc0`, the baseline rows will be missing it. When DuckDB unions
  the lifecourse files across scenarios the missing values become `NULL`,
  which the prevalence/incidence/cost aggregations treat as 0. If you want a
  genuine zero baseline, create the column in *every* scenario (including the
  baseline) — e.g. set it to `0` when the intervention is not active.
* **Avoid the substring `wt` in custom column names.** The ESP-weighted
  variants of some summaries are produced by string-replacing `wt` with
  `wt_esp` in the SQL. A column name containing `wt` could be corrupted by that
  substitution.
* **`_contd` is opt-in.** Remember to add `"contd"` to the `type` argument of
  `export_summaries()`, otherwise the continuous summary is not written.
* **Type matters.** `_prvl` and `_incd` logic compares against integers
  (`> 0`, `= 1`), so make `_prvl` columns integer counters. `_contd` and
  `_costs` should be numeric.
